# The First Homework: An Introduction to Data Analysis

## The First and Last Miles: Visualization

There is an aphorism to the effect that the first and last miles of any data analysis journey is dominated by visualizations.  That's true, or close enough to it.  The first part of a data analytics project - after the data is available to you! - is to explore the data set through summaries, tables, and lots and lots of visualizations.  The last part is to communicate your conclusions and insights to an audience, and this usually involves creating compelling and informative visualizations.

But to be a little more precise, first you have to *get* that data, then analyze it.  In this exercise we'll assume you have access to the data as a `.csv` file (a comma-separated value file), which means that our task is simply to read the data into RStudio.

Create a new code block for R in RStudio and execute the following command:

```{r}
site <- "https://github.com/"
repo <- "robotgames/data-analysis-exercises/"
file <- "raw/main/data/abalone.csv"
df <- read.csv(paste0(site,repo,file))
```

This could have been accomplished with one big line, like
```{r,eval=FALSE}
df <- read.csv("https://github.com/robotgames/data-analysis-exercises/raw/main/data/abalone.csv")
```
but see what happened in the output?  So instead, we worked with chunks of the path to the file and glued them together later.  *Most of the time R simply "knows" when a line continues and so you can just insert line breaks as needed for readability and style.  However, strings act a little differently.  Inserting a line break literally inserts a line break character into the string...which turns out to be bad for a URL.*

You should see that RStudio's Data pane (upper right area of the RStudio app by default) now includes a line with our dataframe name `df` along with some basic information about it (4177 observations of 7 variables).  We can summarize the dataframe:

```{r}
summary(df)
```

And once we have done that we know which variables are numeric and which are not.  Let's make a quick scatterplot of `Shell.weight` versus `Length`.

```{r}
plot(df$Length,df$Shell.weight)
```
And we've gotten started on the first mile of our data analysis with this data set!  You probably have many, many questions about the details in this exercise.  We'll flesh those out over the course of several of the early assignments.

*Summary: In this exercise you recreated the code above to read the data and create a plot.  You may not understand why the commands work the way they do, and that's okay for now.*

```{r,echo=FALSE}
rm(df,file,repo,site)
```

## Aside: The Data Analysis Process

Let's break down the data analysis process into six discrete stages: Inquiry, Ecology, Processing, Analysis, Communication, and Action.

* **Inquiry** - Ask questions.  Many questions start out vague and broad - "How can we save the world?" and gain specificity during an investigation - "How can we improve the crop yield of this particular grain in this particular region?"  As a data analyst, your goal is to make sure that the questions can be addressed measurable data that is accessible.  The inquiry process may involve some initial data gathering and some exploratory data analysis as the questions under investigation are made more specific.

* **Ecology** - Data exists everywhere in the wild, but you may have to capture it and store it in a zoo for later use.  Understanding what data is out there and how to capture it is a useful skill set.  This process is often known as *data ingestion*, but I prefer the term "ecology" to remind you that data exists in the world **in context**, in a certain system, and that can never be ignored.  The "zoo" where you store your data may be highly structured (relational databases made up of linked collections of tables) or not (a big folder full of web pages, texts of emails, and images).

* **Processing** - Before use in an investigation, the data must be extracted from the zoo and processed.  Data might be problematic in many ways: badly formatted, inconsistent units, stored in different tables, suspiciously out of the range of possibility (a human with a recorded height of 60 feet is probably a data entry error!), or just plain missing.  These issues must be addressed before the data is analyzed.

* **Analysis** - Explore, summarize, visualize, and model the data.  The goal here is to develop responses to the driving questions that are supported by the available data.

* **Communication** - Report the results of the analysis to the appropriate audience.  Communicating results with data is a broad skill set, including: technical skills to create impactful visualizations and animations; design skills to create engaging and aesthetically pleasing supporting materials; critical thinking skills to describe how the conclusions were reached; and communication skills to engage the audience.

* **Action** - Decision makers use the conclusions of the analysis to take action.

We'll also break down data contexts - the real-world situations when we call for data to be analyzed - into three broad categories.

* **One-offs** - There is a particular question and a data set that seems relevant.  Analyze!  In the one-off, we *don't* need to worry about how often the analysis must be repeated.  The one-off is common in experimental sciences, in which the researcher designs an experiment to collect data, performs it, analyzes the data, and reports.  The data is still archived and may be used later, but there is no expectation of that.

* **Updates** - There is a particular question that must be answered, but the answer - and the data that leads to it! - changes over time.  For example, a company may be interested in selecting a strategy from among many possible.  It gathers data and periodically analyzes that data...and may shift strategies when the data indicates.  In this case, *data engineering* is crucial.  The new data must be gathered and processed repeatedly, and so we would like an automated system with monitoring tools to do so.  Reports must be generated periodically or even updated live on websites, and again we would like an automated system with monitoring tools to do this.

* **Question pools** - The analyst is not focused on one question only, but on a related collection of questions - a question pool.  For example, a corporation might have the broad question "How can we increase shareholder value?"  That can be sharpened to many more specific questions, and the organization might be interested in answering any or all of them.  In this case, the focus is strongly on archiving large data sets and making them accessible to the analysts who need them.  This is similar to the **Updates** context in the sense that the focus is on the engineering aspects, but here the focus is strongly on creating robust, flexible computer architectures to store and retrieve data.  Computer security is also important; the data must be accessible to the analysts who need it, not accessible to those who don't, editable by those who need to do so and not editable by those who have no need.  *You can think about this as you would any shared cloud doc.  Who can view it?  Who can edit it?  Can those permissions change over time and who has the authority to change them?*

Okay, so in this class, what's our focus?

* We're not going to be concerned about data engineering in this course.  So we won't worry about how data is stored or accessed, and we won't worry about automating those processes.  Data engineering has a great deal of overlap with software engineering and with database design, and I highly recommend taking courses in those areas if you are interested.  Having said that...

* We will use tools that allow us to repeat our analyses with updated data sets.

* We will focus strongly on the Analysis and Communication aspects of the data analysis process, and somewhat less heavily on the Processing and Inquiry aspects.  Because we focus heavily on Analysis and Communication...

* We will use tools that strongly support Analysis and Communication operations.  Thus the choice of the R language and the RStudio platform.

*Summary: In this exercise you read about the data analysis process.  When you submit your assignment, just indicate that you read this exercise.*

## Aside: The Assignment Writeup

In this class I expect you to create a new RMarkdown document for each new assignment.  You may want to create a new project for each one (I certainly recommend that).  When you write your responses to assignment exercises, you should present code in visible code chunks and you should connect your code chunks with relevant discussion that acknowledges there is a reader who will be viewing your document.  You can comment your code with short notes, but that is not a replacement for the connecting language between code chunks.

That is, get in the habit of thinking of each RMarkdown document as a full-featured document, part of which includes code.

*Summary: You read about the assignment expectations.  In your assignment writeup, indicate that you read and understand this exercise.*

## From Math Expressions to R

The syntax for mathematical expressions in R works as in most spreadsheets and calculators.

Create a new RMarkdown document.  Evaluate each of the following mathematical expressions with a chunk of R code.  For example, if the expression is $3e^{2*1.45-1}$ you would write the chunk of R code:

```{r}
3*exp(2*1.45-1)
```

and evaluate to arrive at your answer.  Connect your evaluations with some relevant text, like "this is my answer to part (a)".  *You may have to search to discover how to write some mathematical functions in R.  This is an opportunity to practice searching for technical details about your coding language, an invaluable skill.*

(a) $3\cdot 5^2-2.4\cdot 10^{-1}$  Answer:

```{r,echo=FALSE}
3*5^2-2.4*10^(-1)
```


(b) $\displaystyle \sqrt{14-10\cdot 0.321}$  Answer:

```{r,echo=FALSE}
sqrt(14-10*0.321)
```


(c) $\displaystyle \sqrt{\frac{14+5.2}{2.1-0.9}}$  Answer:

```{r,echo=FALSE}
sqrt( (14+5.2)/(2.1-0.9) )
```


(d) $\displaystyle 0.2\cdot e^{2.4+0.2^2}$ Answer:

```{r}
0.2*exp(2.4+0.2^2)
```


(e) $\displaystyle \ln(0.2)-\ln(\sqrt{3})$  Answer:

```{r, echo=FALSE}
log(0.2) - log(sqrt(3))
```

*Summary: In this exercise you evaluated some mathematical expressions using R like a calculator.  If you're curious about how I formatted the math expressions to look like math expressions ... well, did you do a web search?  Why not?  Try it!  Get in the habit of formulating good search questions and looking for answers to technical questions.*

## Rounding Numbers in R

(a) Execute the following code chunk to see what happens.

```{r}
round(1.234)
round(1.234,1)
round(1.234,2)
```

(b) Execute the following code chunk to see what happens.

```{r}
floor(1.234)
ceiling(1.234)
floor(-1.234)
ceiling(-1.234)
```

(c) Explain in plain language what `round`, `floor`, and `ceiling` do.

*Summary: You experimented with common rounding functions.*

## Scientific Notation in R

(a) Execute the following code chunk to see what happens.

```{r}
7e3
```

(b) Multiply `2.1 x 10^8` and `3.4 x 10^7`.  Answer:

```{r,echo=FALSE}
(2.1e8) * (3.4e7)
```

*Summary: You learned how to use scientific notation in R.*

## Variables in R

(a) Use an R code chunk to store 4 in the variable $x$ and -2 in the variable $y$.
```{r,echo=FALSE}
x <- 4
y <- -2
```

(b) Now use an R code chunk to evaluate $2x^2+3xy-5y^2$.  Answer:
```{r,echo=FALSE}
2*x^2+3*x*y-5*y^2
```

(c) Now, write a single code chunk that does (a), then (b).  Evaluate to make sure you have a correct working chunk.  Then in your code chunk, change the 4 to a 0 and the -2 to a 3 and re-evaluate.  Answer:
```{r,echo=FALSE}
x <- 0
y <- 3
2*x^2+3*x*y-5*y^2
```

(d) Based on your work in (c), what is the benefit of storing data in variables rather than typing data in by hand?
## Characters (Strings) in R

(a) Execute the following code block to define strings and store them in `a` and `b`.  Notice the extra space at the end of the second string.

```{r}
a <- "apple"
b <- 'banana '
```

*Note:* You may use double or single quote to surround a string in R.  Double quotes are my own preference, because single quotes are easily mistaken for the back tick (next to the 1 key on most keyboards).

(b)  Use the `nchar` function to count the letters in `a`.  Answer:

```{r,echo=FALSE}
nchar(a)
```

(c) Use the `trimws` function to trim the extra whitespace (spaces, new lines, etc.) from `b` and store the result in `b`, then use `nchar` to count the characters in the new `b`.  Answer:

```{r,echo=FALSE}
b <- trimws(b)
nchar(b)
```

(d) Join the two strings `a` and `b` with the `paste` function, using `,` as a separator with `sep`.  Answer:

```{r}
paste(a,b,sep=",")
```

(e) Convert the number `1204.73` to a string and store it in `x` by executing the following code chunk.

```{r}
x <- as.character(1204.73)
```

(f) Create a string expressing a distance with units, meters, by pasting together the string `x` from (e) with "meters".  Answer:

```{r,echo=FALSE}
paste(x,"meters")
```


## Logicals in R

A logical data type is usually the result of evaluating an expression as true or false.  In R, these are written `TRUE` and `FALSE`.

(a) Execute the following code chunk to evaluate the truth or falsity of the inequality.

```{r,eval=FALSE}
sqrt(1234) > 36
```
(b) The results of evaluating logical statements can be stored in variables.  Execute the following code chunk to store logical results in `a` and `b`.

```{r}
a <- 3.2^2 < 10
b <- 7.8*15 > 10^2
```

*Note:*  You should be able to see the results in the environment pane in RStudio.

(c) Logical statements can be connected by the "and", `&`, and "or", `|`, logical connectives.  Evaluate the following code chunk to see what happens.

```{r}
a <- 3 < 5
b <- 7 > 10
a & b
a | b
```

(d) The unary "not" operator is denoted `!` in R.  Execute the following code chunk to see what happens.

```{r}
!(3<5)
```

(e) We may encounter situations in which we are given a numeric value `x` and we wish to evaluate whether it meets every one of a set of given constraints.  Write a code chunk with `x <- ` a value at the top, and that evaluates whether `x` is less than 10, whether the square of `x` is greater than 3.1, and whether the cube of `x` is less than 70.

## Data Types in R

State the data type of each of the following expressions in R.  You can use the `class` function, as in

```{r}
class(1.2)
```

and

```{r}
class(2.3>0)
```

*Note:* If you need finer-grained understanding of the internal representation of data, you can use the `typeof` function instead of class.  Generally, `class` will tell us more about data-oriented things and `typeof` will tell us how the data is represented in our computer's memory.

(a) `7.4^2`

(b) `"December 2, 1921"`

(c) `"greed^2"`

(d) `32.1-10 > 4.0`

(e) The class of an expression can be unexpected, until you understand an important fact about R's behavior: when a logical is used where a numeric *should* be used, the logical is often cast to 1 when `True` and 0 when `False`.  Here's an experiment to see this in action.  

Compute `exp(32.1-10 > 4.0)` in an R code chunk and see what happens!  Compare to $e^1$; did you get the same result?  Now try to evaluate `exp(3 > 5)`.  What happens?  And in particular, what is the class of `exp(3 > 5)`?


## Different Assignment Operators

There are usually a number of different ways to "say" something in R, as in most languages.  Instead of asking "how do I say this?", ask "what are the different ways I can say this, and which is best for this context?"

There are a number of different notations for assigning a value to a variable.  In this exercise we'll look at several.

(a) Execute the code chunk and make sure that `2` is printed.  This verifies that `2` is stored in `a`.

```{r}
a <- 2
a
```

(b) Now let's store `3` in `a` using a different notation.  Execute the following code chunk!

```{r}
a = 3
a
```

(c) Next, store `4` in `a`.  Execute the following code chunk.

```{r}
4 -> a
a
```

(d) This is getting a bit ridiculous, but ... let's store `5` in `a`, and then `6` in `a`.  Execute the following code chunk.

```{r}
a <<- 5
a
6 ->> a
a
```

(e) While the `=` sign is most familiar as an assignment operator to coders coming from other languages, it is not the best assignment operator to use in R.  Why not?  First, it is easy to confuse the `=` assignment operator with the `==` logical test.  Second, the arrows are a lot more flexible.

But why do we have single-headed arrows and double-headed arrows?  Execute the following two code chunks.  We'll discuss functions in more detail later.

```{r}
a <- 1
f <- function() {
  a <- 2
  print(a)
}
f()
a
```
*Note:* calling the function `f` assigns `2` to `a` inside the function's space, or *scope*, and then prints whatever is in `a`.  But this does not affect the values stored in variables outside of the scope of the function.  So, when we execute `a` after `f()`, you should see `1` printed, because the value of `a` outside of `f` did not change.

Same code chunk except with a double arrow head inside the function scope!
```{r}
a <- 1
f <- function() {
  a <<- 2
  print(a)
}
f()
a
```

*Note:*  This time, executing the function `f` does result in a change to the variable `a` outside of the function's scope; that's the effect of the double arrow head!

Generally speaking, you should not change the value of a variable outside the scope of the function without an exceptionally good and well-articulated reason.

## Vectors - Not the Physics Kind

(a) Use an R code chunk to write the vector `c(3,2,5,0)` and store it in the variable `x`.
```{r,echo=FALSE}
x <- c(3,2,5,0)
```


(b) Evaluate `x^2` and `exp(x)`.  What happens?  Numerical answer for $x^2$:
```{r,echo=FALSE}
x^2
```


(c) Use an R code chunk to write the vector `c(-2,-1,-4,1)` and store it in the variable `y`.  Then evaluate `x+y`.  What happens?  Numerical answer:
```{r,echo=FALSE}
y <- c(-2,-1,-4,1)
x+y
```

(d) Take a moment and write in plain language how R seems to handle arithmetic with vectors.

(e) Vectors may be non-numerical as well.  Use an R code chunk to store the vector `c("apple","banana","cat")` in the variable `L`.

```{r,echo=FALSE}
L <- c("apple","banana","cat")
```

(f) Elements of vectors must have the same type of data throughout.  Execute the following chunk of code and see whether you understand what R is doing to make sure that all the data in the vector `a` is the same class.

```{r}
a <- c(1.2,"cat",TRUE)
a
```

(g) You may use `c()` to concatenate vectors into a single vector.  Execute the following code chunk to see what happens.

```{r}
a <- c(1,3,5)
b <- c(2,4,6)
together <- c(a,b)
together
```

What would have happened if we replaced `c(a,b)` by `c(b,a)`?

## R Commands for Building Vectors from Patterns

(a) You can create a *range* of values using `:`.  Execute the following chunk of code to see what happens.

```{r}
a <- 1:5
a
b <- -4:7
b
```

(b) You can create a vector of a pattern of values using the `seq` command.  Execute the following code chunk to see what happens.

```{r}
a <- seq(1,13,by=2)
a
b <- seq(-1,1,length.out=20)
b
```

(c) In many cases you can use vectorized arithmetic to create sequences quickly.  Execute the following code chunk to see what happens.

```{r}
a <- seq(3,24,by=3)
b <- 3*(1:8)
a
b
```

(d) You can create a vector with one element repeated many times using `rep`.  Execute the following code chunk to see what happens.

```{r}
a <- rep("red",10)
a
```

*Note:* the `rep` command can be useful for labeling data in data frames.

## Recycling Vectors

R exhibits a peculiar behavior with respect to arithmetic of vectors, called *recycling*.  It is occasionally useful in practice.  More often, it causes code to execute in ways that seem strange unless you are aware of this behavior.

(a) Use an R code chunk to store the vector `1:4` in `a` and `5:8` in `b`.

```{r,echo=FALSE}
a <- 1:4
b <- 5:8
```

(b) Now execute the following code chunk to see what happens when we multiply `a` and `b`.

```{r}
a * b
```

(c) Next, store the vector `c(-1,1)` in `b` and compute `a * b`.  Answer:

```{r}
b <- c(-1,1)
a * b
```

*Note:*  The behavior we are seeing here is called *recycling* in R.  When executing `a * b`, the first two values are multiplied, then the second two, and so on.  If one vector "runs out" of values, then R simply returns to the beginning of that vector for the next value and cycles through again.  This continues until all the values in the longer vector have been used.

(d) Execute the following code chunk to see what happens.

```{r}
1:5 * c(-1,1)
```

*Note:*  R provides a warning when the shorter vector is not a multiple of the longer vector, but it uses the recycling behavior anyway.

## Factors

A special - and incredibly important! - type of vector is a *factor*, also known as a categorical variable.  A factor is a vector whose entries represent categories to which a data entry belongs.  The categories in a factor are known as its *levels*.  A great way to think about factors is as responses to multiple choice questions.

Execute the following code chunk to see what happens.

```{r}
a <- c(rep("A",10),rep("B",5),"C","C")
b <- factor(a)
a
b
```

*Note:* The factor `b` prints the levels as well as the entries.

## Functions Measuring Properties of Vectors 

Some functions in R take vectors as inputs and produce single numbers as outputs.  In this exercise I'd like you to look up the documentation for the `sum` and `length` functions in R.  Then, for the vector `x` from Exercise 4, use R code chunks to compute the sum of `x`, the length of `x`, and the mean of `x`.  Answer for mean of $x$:
```{r,echo=FALSE}
mean(x)
```


## Extracting Elements From Vectors 


(a) Write and evaluate the following R code chunk.  Make sure you understand what it is doing!
```{r}
x <- 3*(1:10)
```

(b) Write a code chunk that extracts the first three elements of `x`.

(c) Write a code chunk that extracts the last three elements of `x`.  *Hint: `tail()` function.*

(d) Write a code chunk that extracts the 1st, 5th, and 7th elements of `x` in one vector.  Answer:
```{r,echo=FALSE}
x[c(1,5,7)]
```


(e) Write a code chunk that attempts to extract the 2nd, 4th, and 13th elements of `x` in one vector.  What goes wrong, and why?





## Extracting Elements With Logicals

Use `x <- 1:1000` to create a vector and store it in `x`.  Then,
```{r,echo=FALSE}
x <- 1:1000
```


(a) Evaluate `x[x^2<10]` in a code chunk.  The output should be much shorter than the original vector `x`.

(b) To figure out what the code in (a) is doing, let's break it down.  The code splits naturally into two pieces.  The *filter* is the logical vector $x^2<10$ and the data vector is $x$.  How many of the filter's entries are true?  Does this help explain why `x[x^2<10]` has so few entries?

(c) Find all of the values in the vector `x` that satisfy the inequality $|x-300|\leq 25$.  Absolute value is a function in R, `abs`.  Answer:
```{r,echo=FALSE}
x[ abs(x-300)<= 25 ]
```


*Note: Filtering with logical vectors is a tremendously important operation.  We'll see it come up again in this homework and throughout the course.*


## Positions or Values? 

In some cases, we don't need to know the *values* in a vector that satisfy some filter, but we need to know their *positions* in the vector.  Note that the position of a value in a vector is often called its *index*.

Use `x <- 2*(1:10)` to create a vector and store it in `x`.  Make sure you know what the vector `x` contains!  Then,
```{r,echo=FALSE}
x <- 2*(1:10)
```


(a) Evaluate `which(x^2<75)` in an R code chunk.  Does this return what you think it ought to?  Answer:
```{r,echo=FALSE}
which(x^2<75)
```


(b) Which indices in `x` have data satisfying the inequality $|x-4|<3$?  Use an R code chunk and the `which` function to determine your answer.  *Remember, the indices are the positions in the vector, not the values.  So your answer here should look like "the third and fifth entries" or something along those lines.*  Answer:
```{r,echo=FALSE}
which( abs(x-4) < 3 )
```


## Extracting With Conditions From Corresponding Vectors

Suppose we have collected data about some children.  For each child, we have measured their age in years and their height in inches.  We have stored this data in two corresponding vectors,

```{r}
age <- c(10,8,5,11,10,6,5)
height <- c(52,48,41,60,54,43,43)
```

By *corresponding*, I mean that the first entry in each vector refers to child 1's measurements, the second entry in each vector refers to child 2's measurements, and so on.

Write these vectors in an R code chunk.  Then,

(a) Your teammate executes the expression `age[height>62]`.  What question are they trying to answer?  How many entries does the result have?

(b) Write an R code chunk to find the ages of all children taller than 50 inches.  Answer:
```{r,echo=FALSE}
age[ height > 50 ]
```


(c) Write an R code chunk to find the heights of all children under 10.  Answer:
```{r,echo=FALSE}
height[ age < 10 ]
```


(d) Let's return to your teammate's code, `age[height>62]`.  How does the code `age[which(height>62)]` compare?  Does it produce the same result?

(e) Hmm.  That last exercise raises a question of why we need the `which` function in R.  Let's compare two ways to filter for children with heights greater than 62 inches.  Evaluate the expressions `height>62` and `which(height>62)`.  Clearly one is shorter to type, but what are the other differences?  In particular, would one be easier to store in a file than the other?

(f) The `which` function has some related functions, `which.max` and `which.min`.  Read the R documentation for these functions, and then use them to find the indices of the tallest and shortest children in the data.  Answer for tallest:
```{r,echo = FALSE}
which.max(height)
```


(g) Use your results from part (f) to find the ages of the tallest and shortest children in the data set.  Answer, age of tallest:

```{r,echo=FALSE}
age[which.max(height)]
```


*Note:* We could have answered any of these questions by a simple inspection of the data set, because it has very few records.  In order to learn the software, make sure that you are always performing your work as if the data has millions of records.  Do your results rely at any point on a visual inspection?  Can you work around that?  These are questions we will ask ourselves throughout the course.

## Re-leveling Factors

Regardless of the method used to subset, factor variables sometimes pose a problem.

(a)  Execute the following code chunk to see what happens.

```{r}
a <- factor(c("A","A","B","B","C"))
b <- a[1:4]
b
```

(b) In part (a), what are the distinct values in `b`?  What does R state as the levels of `b`?

(c) In some applications we will need to retain the "memory" of the levels from which the subset was taken.  That is, we might need to know that `b` was drawn from a factor with levels `levels(b)`.  This is the reason that R preserves this information by default.

But, in many applications, we simply need to know the levels of `b`.  We can re-level `b` as in the following code chunk.

```{r}
b <- factor(b)
b
```

(d) In some cases we might need to specify an order on the levels of the factors.  Execute the following code chunk to see what happens.

```{r}
a <- factor(c("A","A","B","B","C"),
            levels=c("A","B","C"),
            ordered=TRUE)
a
a <- factor(c("A","A","B","B","C"),
            levels=c("B","A","C"),
            ordered=TRUE)
a
```

*Note:*  Ordering levels in a factor can be useful in a variety of situations, such as ordering the months in a year or the days in a week.  Ordering levels helps other functions like plotters create structures that are aware of the intended order.

(e) In some cases we may want to rename the levels of a factor.  Execute the following code chunk to see what happens.

```{r}
a <- factor(c("Level A1","Level B1","Level C1"))
a
levels(a) <- c("A","B","C")
a
```


## Tables

(a) Execute the following code chunk to see what happens.

```{r}
a <- c(rep("a",10),rep("b",30),rep("c",5))
table(a)
```

(b) Execute the following code chunk to see what happens.

```{r}
a <- c(rep("a",10),rep("b",30),rep("c",5))
b <- c(rep("a",20),rep("b",5),rep("c",20))
table(a,b)
```

## Lists

Vectors in R must contain data that is all of the same class.  Knowing that all of the data in a container is the same class is useful for coders, who can then write functions assuming the data is all numerical, or all character, or all logical.  But, sometimes we will need containers that are able to hold data of different classes.  In R, this is the *list*.

(a) Execute the following code chunk to build a list `a`.

```{r}
a <- list("apple",1204.73,FALSE)
```

(b)  Extract the second element of `a` using `a[[2]]`.

(c) Execute the following code chunk to build a named list `b`.

```{r}
b <- list(dog="Fido",size=1.2,color="reddish")
```

(d) Execute the following code chunk to see what happens.

```{r}
b[[1]]
b[["size"]]
b$color
```

*Note:* This gives a few different ways of accessing the elements of a list.  The last method using `$` will be used often when we move to *data frames*, a special kind of list.

(e)  Execute the following code chunk to see what happens.  The object returned is a list.

```{r,eval=FALSE}
b[1:2]
```

(f) Execute the following code chunk to see what happens.

```{r}
b[2:3][[1]]
```

