# Day 16: DW - Joining and Reshaping

## Joining data sets

We often wish to join data sets from different sources.  The `tidyverse` has database tools for accomplishing this.  In this exercise, you'll build small synthetic data sets and use the database tools to join them together.

(a) Execute the following code chunk to build and display two data sets.

```{r}
left <- tibble(x=runif(4),y=runif(4),
                group=c("a","a","a","b"))
right <- tibble(group=c("a","b","c","d"),z=c(10,5,1,0))
print(left)
print(right)
```

(b) For the data sets defined in (a), think of `left` as the "main" data set.  Data set `right` is a tbale that has some additional information about the groups listed in the variable `group`.  We want to copy that information into each row in `left` corresponding to the group in the row.  The function `left_join` will accomplish this, holding `left` constant and attaching information from `right`.  Execute the following code block to see what happens.  Did it work?

```{r,eval=FALSE}
left_join(left,right,by="group")
```

*Note:*  The `right_join` function is analogous to `left_join`.

(c) Go back to (a) and redefine `left` to have an additional `x` value, an additional `y` value, and one extra value in group, `e`.  Then run the `left_join` code chunk in part (b).  What happens?

(d) Execute the following code chunk to build and display two data sets.

```{r}
left <- tibble(x=runif(4),y=runif(4),
                group=c("a","b","d","e"))
right <- tibble(group=c("a","b","c","d"),z=c(10,5,1,0))
print(left)
print(right)
```

Then, execute the following code chunk to see what happens.

```{r,eval=FALSE}
inner_join(left,right,by="group")
```

What rows were omitted and why?

(e)  Rerun the second code chunk in part (d) but replace `inner_join` with `full_join`.  What happens?

(f) A common operation is to join data on observational units to another set of data aggregated by groups of those observational units.  

For example, we might have a set of data on vote counts by reporting precinct in the U.S., and we might have a second set of data gathered at the county level on economic variables such as median income and employment status.  Having an interest in understanding why people vote as they do, we might want to aggregate the first set of data to the county level by summing votes, and then join the two data sets by county.  

In this exercise you'll deal with exactly this sort of problem.

First, let's import a data set.

```{r}

```


## Pivoting longer
An observational unit in a data frame is a collection of keys that uniquely identify the unit.  For example, there is a difference between the observational unit "Barry Allen" and the observational unit "Barry Allen at age 27".

Unfortunately, the keys for the observational unit can be treated as features.  This causes all sorts of confusion, and in particular, can sometimes cause data to be entered in ways that are less amenable to analysis.  Here's an example.

Suppose we have collected height data on children over time.  Here are our records for three children.

```{r}
children <- c("Barry","Jay","Wally")
age_5 <- c(42,46,43)
age_10 <- c(62,67,62)
age_15 <- c(68,72,67)
df <- data.frame(Name=children,
                 age_5=age_5,
                 age_10=age_10,
                 age_15=age_15)
```

In this data set the observational units are children and the data is height changing over time.  But right now, the variable associated with time - age - is trapped in the column names, and is not really part of our dat aset.  We may prefer to have this data in a table with three variables: name, age, and height.  We'll use the `pivot_longer` function.

(a) Execute the following code chunk to see what happens.  Make sure to read the code and check the documentation carefully to understand what is happening at each step.

```{r,eval=FALSE}
df %>%
  pivot_longer(cols=2:4,names_to="age",values_to="height") %>%
  mutate(age=str_split(age,"_") %>% map_chr(.,2)) %>%
  mutate(age=as.numeric(age))
```

(b) Use the code chunk in part (a), store the data frame created in the variable `df2`, and then create and display a new data frame computing the mean height and standard deviation in height at each `age` (use `group_by` and `summarize`).  Would this have been easy to do with the original data frame `df`?

(c) Using `df2` as created in (b), summarize the height measurements by `Name`.  Would this have been easy to do with the original data frame `df`?  Would it have been easy to do if there were forty ages in the original data frame rather than three?

(d) Use `pivot_longer` to write all the numeric measurements in the `iris` data set in one column, `measurement`.

*Note:* Pivoting wider is the "inverse" operation to pivoting longer, and is used to change focus on the observational unit.