# Day 25: More Hypothesis Tests

ANOVA, Chi-square, comparing distributions with K-S tests or Wasserstein distance

## ANOVA and Tukey's Honest Significant Difference

Load the `ToothGrowth` data set (in R).  

(a) Examine the `dose` variable.  Should it be a numeric or a factor?  Why?

(b) Use an ANOVA test to determine whether there is a difference among the groups defined by different dose levels.  Answer: Yes, with a p-value much less than 0.
```{r,echo=FALSE,eval=FALSE}
data("ToothGrowth")
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
my.aov <- aov(len ~ dose, data=ToothGrowth)
summary(my.aov)
```

(c) Use `TukeyHSD` to analyze the differences in the pairs of groups in this data set.

## Exploring ANOVA and TukeyHSD

Let's create some synthetic data to see how ANOVA performs.

(a) Use the following code to create synthetic data.  Describe what this is doing.

```{r}
set.seed(12345)
adata <- rnorm(15,mean=100,sd=10)
bdata <- rnorm(30, mean=102,sd=10)
cdata <- rnorm(50,mean=110,sd=8)
the_labels <- c(rep("a",length(adata)),
                rep("b",length(bdata)),
                rep("c",length(cdata)))
df <- data.frame(x=c(adata,bdata,cdata),
                 y=factor(the_labels))
```

(b) Perform an ANOVA test (with a 5% rejection level) on this synthetic data to determine whether there is a significant difference among the means of the three groups.  Answer:  There is, and the summary of the ANOVA output is...
```{r,echo=FALSE}
my.aov <- aov(x ~ y,data=df)
summary(my.aov)
```

(c) Use `TukeyHSD` (at a 5% rejection level) to test whether there are differences in the means of pairs of groups.  Answer: There are significant differences between a and c and between b and c, but not between a and b, and here is the output...
```{r,echo=FALSE}
TukeyHSD(my.aov)
```

(d) Return to part (a).  Rebuild your data set in exactly the same way (using the same `set.seed(12345)` statement), except use a standard deviation of 30 for group c.  Perform a `t.test` to compare groups a and c.  Does the `t.test` find a significant difference in their means?  Answer: Yes, and here is the output...

```{r,echo=FALSE}
set.seed(12345)
adata <- rnorm(15,mean=100,sd=10)
bdata <- rnorm(30, mean=102,sd=10)
cdata <- rnorm(50,mean=110,sd=30)
the_labels <- c(rep("a",length(adata)),
                rep("b",length(bdata)),
                rep("c",length(cdata)))
df <- data.frame(x=c(adata,bdata,cdata),
                 y=factor(the_labels))
t.test(adata,cdata)
```

(e) Perform your ANOVA and `TukeyHSD` analyses again, this time for the data set you constructed in part (d).  What happens?  Is this difficult to interpret?  Does it agree with your `t.test` results in part (d)?  Answer:  Here's what happens!  ANOVA finds that there is a significant difference among the sample means of the groups at the 5% level.  However, TukeyHSD finds no significant difference between any pairs of groups at the 5% level.  As it turns out, individual t.test applications to pairs of groups can inflate the size of the difference between two groups when compared to the larger context of all of the data.  Tukey proposed a different way of measuring the difference between groups in a larger data set, which he called his Honest Significant Difference, or HSD...and that's really where the name TukeyHSD comes from!  Google "Tukey Honest Significant Difference" to learn more!

```{r,echo=FALSE}
set.seed(12345)
adata <- rnorm(15,mean=100,sd=10)
bdata <- rnorm(30, mean=102,sd=10)
cdata <- rnorm(50,mean=110,sd=30)
the_labels <- c(rep("a",length(adata)),
                rep("b",length(bdata)),
                rep("c",length(cdata)))
df <- data.frame(x=c(adata,bdata,cdata),
                 y=factor(the_labels))
my.aov <- aov(x ~ y,data=df)
summary(my.aov)
TukeyHSD(my.aov)
```



