# Day 26: Linear Models and Regression

## Reading lm Output (more Gordon exercises?)

Gordon loaded some data and created a model with the following code:

```{r}
df <- read.csv('https://github.com/cbrown-clu/Math-331-Data-Analysis/raw/master/B.%20Data%20sets/abalone.csv')
df <- df[-which(df$Sex=="M"),]
df$Sex <- factor(df$Sex)
diam <- lm(Shell.weight ~ Diameter,data=df)
summary(diam)
```

Describe the model that Gordon created.  Using the model output, address the question of whether this is a good model.


## Using lm to Model Quadratic Relationships


(a) Load the abalone data set that Gordon used in Exercise 1.  Summarize it, and convert variables to factors where reasonable.

(b) Plot the `Shell.weight` versus the `Diameter`.

```{r,echo=FALSE,eval=FALSE}
ggplot(data=df,aes(x=Diameter,y=Shell.weight)) +
  geom_point()
```

(c) Based on your plot in part (b), it should be apparent that a linear relationship is not quite correct.  Create a linear model for a quadratic relationship of `Shell.weight` versus the `Diameter`.

```{r,echo=FALSE,eval=FALSE}
mod <- lm(Shell.weight ~ Diameter + I(Diameter^2),data=df)
summary(mod)
```


(d) Is the quadratic relationship you created in part (c) better than the linear relationship from exercise 1?  Explain.

## Comparing Two Models Created With lm (should add notes here on model comparison)


Using the abalone data set, create two different linear models.  Use `Shell.weight` as the response variable in each.  In one, use `Length` as the predictor variable and use a quadratic relationship.  In the other, use `Height` as the predictor variable and use a quadratic relationship.  Compare the two models.


## Overfitting with lm (need to tweak this example)


Before returning to the abalone data set, let's explore an issue in model creation.

(a) Start by creating a synthetic data set using the following code:

```{r}
set.seed(12345)
x <- runif(500,-1,1)
noise <- rnorm(length(x),mean=0,sd=0.20)
y <- 1 - x^2 + noise
syn <- data.frame(x=x,y=y)
```

(b) Plot `y` versus `x` in the data set you created in part (a).  Answer:

```{r,echo=FALSE}
library(ggplot2)
ggplot(data=syn,aes(x=x,y=y)) + geom_point()
```

(c) Fit a quadratic relationship for `y` versus `x`.  What is the $R^2$ value?  Answer:

```{r,echo=FALSE}
syn.lm <- lm(y ~ x + I(x^2),data=syn)
summary(syn.lm)$r.squared
```

(d) Now fit a cubic relationship for `y` versus `x`.  What is the $R^2$ value, and is it (slightly) better than with the quadratic model?  Answer:
```{r,echo=FALSE}
syn.lm <- lm(y ~ x + I(x^2)+I(x^3),data=syn)
summary(syn.lm)$r.squared
```

(e) Try fitting higher degree relationships for `y` versus `x` and compare the $R^2$ values.  Do they continue to increase (slightly)?

*Note: One issue that we will sometimes encounter is an optimizer mindset.  When we encounter a simple measurement like r squared, it is tempting to adjust the model to continue to make small gains.  This is usually not a great idea.*


