# Day 22: Hypothesis Tests

## Hypothesis Testing - Single Sample Mean


Let's imagine that we gather a sample of 50 irises of the setosa species, we measure their petal lengths, and we compute the mean of those petal lengths.  You can find our measurements in the `iris` data set.

(a) Perform a formal test of hypothesis testing the hypothesis that the population mean petal length of setosa irises is 1.6.  Use a two tailed test and a confidence level of 95%. Make sure to state your p-value explicitly.

(b) Use the output of `t.test` to determine which hypothesized values for the population mean petal length will be rejected by a test with confidence level 95% (rejection level 5%).


## Hypothesis Testing - Difference in Population Means


(a) Make a plot comparing the petal lengths by species in the iris data set.  Are the means for the versicolor and virginica species significantly different, based on your graph?

(b) Use a t-test at the 95% confidence level to test the hypothesis that the mean petal lengths of the versicolor and virginica species are the same.  (Note: unpaired and two-tailed!)


## Estimating the Power of a Hypothesis Test (check answer in part a)

Let's suppose you collect the following data (copy this into your RMarkdown document and execute it).

```{r}
set.seed(1234)
data <- rgeom(40,prob=1/6)
```

(a) Estimate the 95% confidence interval on the population mean based on this data.  You can do this with a simulation, but `t.test` may give a faster method.  Answer:

```{r,echo=FALSE}
test_info <- t.test(data,mu=2)
test_info$conf.int
```

(b)  Suppose that the true mean of the population is 4.8. (call this hypothesis H1.)  What is the power of the test in rejecting the null hypothesis H0: population mean is 3.0, at a rejection level of 5%, with a two-tailed test?  Answer:

```{r,echo=FALSE}
error <- sd(data)/sqrt(length(data))
tickmarks <- qnorm(c(0.025,0.975),
                   mean=3.0,sd=error)
1 - (pnorm(tickmarks[2],mean=4.8,sd=error) - 
  pnorm(tickmarks[1],mean=4.8,sd=error) )
  
```

(c) Suppose that the true mean of the population is 3.8. (call this hypothesis H1.)  What is the power of the test in rejecting the null hypothesis H0: population mean is 3.0, at a rejection level of 5%, with a two-tailed test?  Answer:


```{r,echo=FALSE}
error <- sd(data)/sqrt(length(data))
tickmarks <- qnorm(c(0.025,0.975),
                   mean=3.0,sd=error)
1 - (pnorm(tickmarks[2],mean=3.8,sd=error) - 
  pnorm(tickmarks[1],mean=3.8,sd=error) )
```

(d) Why the difference in power?  Discuss.

