# Day 34: Clusters with k-Means

## k Means Algorithm for Clustering

In this exercise we'll use the k-means algorithm to cluster some synthetic data.

(a) Execute the following code block to create some synthetic data.
```{r,echo=TRUE}
set.seed(1234)
x <- c(rnorm(100,mean=-2),rnorm(150,mean=2))
y <- c(rnorm(100,mean=-2),rnorm(150,mean=2))
df <- data.frame(x=x,y=y)
```

(b) Plot the synthetic data from (a).  Is it clear that there are two clusters?  Answer:
```{r}
library(ggplot2)
ggplot(data=df,aes(x=x,y=y)) +geom_point()
```

(c) Use `kmeans` to create a model of the clusters.  Use the option `centers=2`.
```{r}
km.model <- kmeans(df,centers=2)
```

(d) Attach the cluster information to the data frame `df`.  My code looks like the following; yours may differ depending on the names you chose for your variables.
```{r,echo=TRUE}
df$cluster <- factor(km.model$cluster)
```

(e) Now let's plot the data again, but this time color the points by cluster.  Answer:
```{r}
ggplot(data=df,aes(x=x,y=y,col=cluster)) + geom_point()
```

(f) Use the `cluster` library to visualize the clusters.  My code:
```{r,echo=TRUE}
library(cluster)
clusplot(df[,1:2],km.model$cluster)
```

*Note that clusplot transforms the data somewhat, so the data points in the two plots may look numerically different, but qualitatively similar.*
